---
title: "PS4"
author: "Jeffery Zhang"
class: "INFO 201"
TA: "Xinyu Chang (AD)"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## ---------- 1: Working Directory ----------

# 1.1
The working directory of the RStudio console is at C:/Users/xuz-t/OneDrive/Documents/INFO_201_Scripts.

# 1.2

The .rmd is saved to C:/Users/xuz-t/OneDrive/Documents/INFO_201_Scripts.

# 1.3
They are the same. This matters because it helps streamline file access.

## ---------- 2: Create Your Own Data ----------

# 2.1
```{r name_vector}
names <- c("Allan", "Bill", "Charlie", "David", "Ethan")
print(names)
```

This is a vector of five students.

# 2.2
```{r math_sample}
math <- sample(100, 5, replace = TRUE)
print(math)
```

Random sample of math grades, from 0-100.

# 2.3
```{r japanese_sample}
japanese <- sample(100, 5, replace = TRUE)
print(japanese)
```

Random sample of japanese grades, from 0-100.

# 2.4
```{r dance_sample}
dance <- sample(100, 5, replace = TRUE)
print(dance)
```

Random sample of dance grades, from 0-100.

# 2.5
```{r grades_dataframe}
grades <- data.frame(Name = names, Math = math, Japanese = japanese, Dance = dance)
print(grades)
```

This is the representative dataframe, constructed out of names and the grades from math, japanese, and dance.

# 2.6
```{r num_students}
num_students <- nrow(grades)
cat(num_students)
```

This is the number of students, or the number of rows in the dataframe.

# 2.7
```{r num_courses}
num_courses <- ncol(grades) - 1
cat(num_courses)
```

This is the number of courses, or the number of columns (minus 1) in the dataframe.

# 2.8
```{r last_two_rows}
last_two_rows <- tail(grades, 2)
print(last_two_rows)
```

This is the last two rows of the dataframe.

# 2.9
```{r create_csv}
library(readr)
data_directory <- "data/"
file_path <- file.path(data_directory, "grades.csv")
write_csv(grades, file_path)
```

This code writes the .csv file to the given filepath.

# 2.10
```{r confirm}
list.files(data_directory)
```

This is a confirmation that the .csv file does, in fact, exist.

## ---------- 3: Indirect Variable Names ----------

# 3.1
```{r indirect_dance}
col <- "Dance"
cat(col)
```

This creates an indirect reference to the dance column within the dataframe.

# 3.2
```{r use_indirect}
grades_list <- grades[, col]
cat("grades: ", grades_list)
```

This is how we access grades for the dance class without explicitly using "dance".

# 3.3
```{r indirect_japanese}
col <- "Japanese"
best_grade <- max(grades[, col])
cat("Best grade in ", col, ": ", best_grade)
```

This is how we indirectly find the best grade for the japanese class without explicitly using "japanese".

# 3.4
```{r average_grade}
for (name in names(grades)) {
  avg_value <- mean(grades[[name]])
  if (!is.na(avg_value)) {
    cat("class ", name, "has an average grade of ", avg_value, "\n") 
  }
}
```

These are all the average grades per class.

## ---------- 4: Data manipulations ----------

# 4.1
```{r GPA}
grades$gpa <- (1/3) * (grades$Math + grades$Japanese + grades$Dance)
gpa_data <- grades[["gpa"]]
print(gpa_data)
```

These are the (pseudo-) GPAs of each student, based on their average grade.

# 4.2
```{r best_student}
best_student_index <- which.max(grades$gpa)
best_student_data <- grades[best_student_index, ]
print(best_student_data)
```

This is the best student, based on their GPA.

# 4.3
```{r best_student_name}
best_student_name <- best_student_data$Name
cat(best_student_name)
```

This is the name of the best student.

# 4.4
```{print_data}
print(grades)
```

Yes, it is right.

# 4.5
```{r is_better_at_math}
grades$better_at_math <- grades$Math > grades$Japanese
print(head(grades, 3))
```

These are the first three lines, representing whether or not the student is better at math than in Japanese.

# 4.6
```{r how_many_better}
num_students_better_math <- sum(grades$better_at_math)
cat(num_students_better_math)
```

This is the number of students who are better at math than at Japanese.

## ---------- 5: Life expectancy ----------

## ---------- 5.1: Load Data ----------

# 5.1.1
```{r createDataframe}
life_expectancies <- read_delim("C:/Users/xuz-t/Documents/INFO_201/PS4/data/life-expectancy.csv")
```

This is RStudio reading the life-expectancy file, with the Tab separator.

# 5.1.2
```{r dataframe_dimens}
num_row <- nrow(life_expectancies)
num_col <- ncol(life_expectancies)
cat(num_row)
cat(num_col)
```

These are the number of rows and columns in the life expectancy data, respectively.

# 5.1.3
```{r var_names}
cat(names(life_expectancies))
```

These are all the variable names within the life_expectancy dataframe.

# 5.1.4
```{r first_five_lines}
head(life_expectancies, 5)
```

These are the first five lines of the life_expectancy dataframe. It generally looks pretty good, but there are a couple of NA entries.

# 5.1.5
Each row in the data set represents a country entry.

# 5.1.6
```{r num_missing_per_var}
missing <- numeric(length(life_expectancies))
for (i in 1 : num_col) {
  missing[i] <- sum(is.na(life_expectancies[[i]]))
  cat("var name: ", names(life_expectancies)[i], "missing vals:", missing[i], "\n")
}
```

These are the number of entries missing per variable. There are a couple of missing life expectancy values, but definitely the variables with the most missing are the GDP per capita values.

## ---------- 5.2: Analyze Data ----------

# 5.2.1
```{r life_expectancy_growth}
life_expectancies$expectancy_growth <- life_expectancies$le2019 - life_expectancies$le1960
head(life_expectancies, 5)
```

This code creates a new variable which shows the growth in life expectancy per country.

# 5.2.2
```{r average_improvement}
average_improvement <- mean(life_expectancies$expectancy_growth, na.rm = TRUE)
cat(average_improvement)
```

This is the average improvement in life_expectancy for every country.

# 5.2.3
```{r min_max_gains_countries}
max_gain_country <- life_expectancies[which.max(life_expectancies$expectancy_growth), "name"]
min_gain_country <- life_expectancies[which.min(life_expectancies$expectancy_growth), "name"]
print(max_gain_country)
print(min_gain_country)
```

These are the countries that gained the most/least life expectancy within the same timeframe.

# 5.2.4
```{r countries_less_than_five}
countries_less_than_5_years <- sum(life_expectancies$expectancy_growth < 5, na.rm = TRUE)
cat(countries_less_than_5_years)
```

This is the number of countries that gained less than 5 years of life expectancy over the timeframe.

# 5.2.5
```{r countries_decreased}
countries_with_decreased_LE <- sum(life_expectancies$expectancy_growth < 0, na.rm = TRUE)
cat(countries_with_decreased_LE)
```

This is the number of countries that saw decreases in life expectancy during the timeframe.